---
title: Predicting the Number of Expected Boating Trips to Lake Somerville using Generalized
  Linear Models
author: "TJ Smith"
output:
  word_document: default
  html_document: default
---

# Import needed packages
```{r,include=FALSE}
library(tidyverse)
library(GLMsData)
library(MASS)
library(gridExtra)
library(statmod)
```

# Loading in data set
```{r,include=FALSE}
rec <- read.csv("rec-trips.csv")
```

# Calculating proportions for target variable values (limited from 0-25)
# Put proportions into a table
```{r,include=FALSE}
tbl <- table(rec$TRIPS)
pp <- tbl / sum(tbl)
dftrips <- data.frame(TRIPS = as.numeric(names(tbl)),
                 freq.TRIPS = as.numeric(pp))
names(dftrips) <- c("Trips", "freq.Trips")
pp <- pp[as.character(0:25)]
round(pp, 4)
```

```{r}
pp <- pp[as.character(0)]
round(pp, 4)
```

# Graphing number of trips to Lake Somerville vs proportion of data
```{r}
p <- ggplot(data=dftrips,mapping=aes(x=Trips,
                                y=freq.Trips))+xlim(0,25)
p <- p + geom_line(col = "gray")
p <- p + geom_point()
p <- p + labs(x = "Number of Trips",
              y = "Proportion")
p
```

#Graph summary of the SO variable
```{r}
p <- ggplot(data = rec,
            mapping = aes(x = as.factor(SO),
                          y = TRIPS)) +ylim(0,15)
p <- p + geom_boxplot()
p <- p + labs(x = "SO",
              y = "Trips",
              title = "")
p
```

# Distribution of the C3 variable compared to Target variable
```{r}
p <- ggplot(data = rec,
            mapping = aes(x = C3,
                          y = TRIPS)) +xlim(0,100)
p <- p + geom_point(pch = 1)
p <- p + geom_smooth(se=FALSE)
p <- p + labs(x = "Dollar expenditure when visiting Lake Somerville",
              y = "Number of Trips")
p
```

# Splitting SO variable into seperate variables based on their value
```{r,include=FALSE}
rec$SO0 = ifelse(rec$SO == "0",1,0)
rec$SO1 = ifelse(rec$SO == "1",1,0)
rec$SO2 = ifelse(rec$SO == "2",1,0)
rec$SO3 = ifelse(rec$SO == "3",1,0)
rec$SO4 = ifelse(rec$SO == "4",1,0)
rec$SO5 = ifelse(rec$SO == "5",1,0)
```

# Final Poisson model
```{r}
mi = glm(TRIPS ~ -1+SO0+SO1+SO2+SO3+SO4+SO5+C3,
         data=rec,family=poisson(link="log"))
summary(mi)
deviance(mi) / df.residual(mi)
```

# Final negative binomial model
```{r}
m0 = glm.nb(TRIPS ~ -1+SO0+SO1+SO2+SO3+SO4+SO5+C3+I+FC3+SKI,
         data=rec)
summary(m0)
```

# Calculating necesarry columns/values for model analysis
```{r,include=FALSE}
rec$m0.mu <- predict(m0, type = "response")
rec$m0.eta <- predict(m0, type = "link")
rec$m0.rW <- resid(m0, type = "working")
rec$m0.wR <- rec$m0.eta + rec$m0.rW
rec$m0.rD <- resid(m0, type = "deviance")
rec$m0.rQ <- qresid(m0)
```

# Overall model accuracy predictor
```{r}
p <- ggplot(data = rec,
            mapping = aes(x = m0.mu,
                          y = m0.rQ))
p <- p + geom_point() + geom_smooth(se = TRUE)
p <- p + labs(x = "Fitted Values",
              y = "Quantile Residuals",
              title = "Negative Binomial Model")
p
```

# Out-of-model variable analysis
```{r}
p <- ggplot(data = rec,
            mapping = aes(x = C1,
                          y = m0.rQ))
p <- p + geom_point() + geom_smooth(se = TRUE)
p <- p + labs(x = "C1",
              y = "Quantile Residuals",
              title = "Negative Binomial Model")
p
p <- ggplot(data = rec,
            mapping = aes(x = C4,
                          y = m0.rQ))
p <- p + geom_point() + geom_smooth(se = TRUE)
p <- p + labs(x = "C4",
              y = "Quantile Residuals",
              title = "Negative Binomial Model")
p
```

# In-model variable analysis
```{r}
p <- ggplot(data = rec,
            mapping = aes(x = C3,
                          y = m0.rQ))
p <- p + geom_point() + geom_smooth(se = TRUE)
p <- p + labs(x = "C3",
              y = "Quantile Residuals",
              title = "Negative Binomial Model")
p
p <- ggplot(data = rec,
            mapping = aes(x = as.factor(SO),
                          y = m0.rQ))
p <- p + geom_boxplot()
p <- p + labs(x = "SO",
              y = "Quantile Residuals",
              title = "Negative Binomial Model")
p
```

# Link function accuracy test
```{r}
p <- ggplot(data = rec,
            mapping = aes(x = m0.eta,
                          y = m0.wR))
p <- p + geom_point() + geom_smooth(se = TRUE)
p <- p + labs(x = "Linear Predictor",
              y = "Working Responses",
              title = "Negative Binomial Model")
p
p <- ggplot(data = rec,
            mapping = aes(x = m0.eta,
                          y = m0.wR))
p <- p + geom_point() + geom_smooth(se = TRUE)
p <- p + labs(x = "Linear Predictor",
              y = "Working Responses",
              title = "Negative Binomial Model")
p
```

# Function Family test (Is negative binomial the right choice?)
```{r}
pt <- ggplot(data = rec,
            mapping = aes(sample = m0.rQ))
pt <- pt + geom_qq(pch = 1) + geom_qq_line(color = "blue")
pt <- pt + labs(x = "Theoretical Quantiles",
              y = "Sample Quantiles")
pt
```

# Final numerical model
$$
log(\mu) = -2.355(SO0) + 1.482(SO1)+2.306(SO2)+1.90(SO3)+2.413(SO4)+2.211(SO5)-.008(C3)-.008(I)+1.040(FC3)+.606(SKI)
$$

# Calculating proportions for target variable values (limited from 0-25)
# Put proportions into a table
```{r,include=FALSE}
tbl <- table(rec$TRIPS)
pp <- tbl / sum(tbl)
dftrips <- data.frame(TRIPS = as.numeric(names(tbl)),
                 freq.TRIPS = as.numeric(pp))
names(dftrips) <- c("Trips", "freq.Trips")
pp <- pp[as.character(0:25)]
round(pp, 4)
```

# Poisson function probability calculation
```{r,md-def-poisson-prob, include=FALSE}
f <- function(mu, k){
  prob <- exp(-mu)*mu^k/factorial(k)
  ans <- mean(prob)
  return(ans)
}
```

# Calculating expected probaility from Poisson model
```{r,md-pm.1-calc-expected-probs, include=FALSE}
rec$mi.mu <- predict(mi, type = "response")
pr <- apply(data.frame(TRIPS = 0:25),
            1, function(k) f(rec$mi.mu, k))
tbl <- data.frame("TRIPS" = 0:25,
                  "Actual" = as.numeric(pp), 
                  "Expected" = as.numeric(pr))
round(tbl, 3)
```

# Calculating expected probaility from neg. binomial model
```{r,md-nb1-calc-expected-probs, include=FALSE}
tmp <- apply(data.frame("TRIPS" = 0:25),
      1, function(k) mean(dnbinom(k, mu = rec$m0.mu, size = 1)))
tbl$Exp.NB <- tmp
round(tbl,3)
```

# Plot showing the actual data vs the negative binomial and Poisson model
```{r,md-plot-probs-both-models}
p <- ggplot(data = tbl,
            mapping = aes(x = TRIPS,
                          y = Actual))
p <- p + geom_line() + geom_point()
p <- p + geom_line(aes(y = Expected),
                   color = "red") + 
  geom_point(aes(y = Expected),
             color = "red")
p <- p + geom_line(aes(y = Exp.NB),
                   color = "blue") + 
  geom_point(aes(y = Exp.NB),
             color = "blue")
p <- p + annotate("text", x = c(20,20,20), y = c(0.6, 0.5,.4),
                  label = c("Actual", "Poisson","Negative Binomial"),
                  color = c("black", "red","blue"))
p <- p + labs(x = "Number of Trips",
          y = "Probability")
p
```

